---
title: "R tutorial"
author: "Jeanette Mumford"
output:
  html_document:
    keep_md: yes
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

## Installing R and getting started
[R](https://www.r-project.org/) is free and I highly recommend using R through the IDE, [RStudio](https://www.rstudio.com/), which offers a lot of helpful functionality like generating R markdown files easily and integrating with github.  A common confusion is updating RStudio is \emph{not} also updating R as they are separate programs.

To install R and RStudio, start by installing R and then install RStudio second.  RStudio should automatically find R when it installs.

Like Python, R requires installing libraries for more advanced tools, but  the installation is typically very easy.  Most libraries you will want to use are located in the [CRAN](https://cran.r-project.org/) (Comprehensive R Archive Network), which can be accessed through R, directly.  The following illustrates how you'd install and load a package.


```{r}
# <-  The comment in R is the hash symbol

# To grab libraries for installation, R will ask you to
# select a CRAN "mirror".  The following line will set the
# mirror permanently so R doesn't ask you over and over again.
#  All the mirrors are listed here: https://cran.r-project.org/mirrors.html

# You only need to run this once ever (not once per session)
options(repos=structure(c(CRAN="http://cran.us.r-project.org")))


# To install a library use the following, note double quotes are used.
# A "newer" feature is dependencies (libraries the library of 
# interest needs) will automatically be installed.

# You only need to run this once ever (not once per session)
#install.packages("ggplot2")

# To load the library so you can access the functions 

# Run once per session if you'd like to use it
library("ggplot2")
```

For the above you will likely get more output than I did when you run install.packages, as I already had it installed.

Now open a new editor (in RStudio click the white page icon in the upper left and select, "R Script", and we'll write a quick "hello world".  Enter the following in the file and save it in a file called "first_script.R".
```{r, results = "hide"}
print("Hello world")
```

I feel it is poor coding form to change working directories, so I always recommend typing out the full paths to files.  So edit the following to reflect where you just stored your file.  This will run your script.
```{r}
source("/Users/jeanettemumford/Documents/Research/Teaching/neurohackweek2017/first_script.R")
```

If you want to run the script from the LINUX prompt you would use "Rscript first_script.R".  

In RStudio, you can simply hit cmd+return (on a Mac) to run the current line that the cursor is on, or you can copy and paste into the R terminal window.  Give that a shot with your single line in first_script.R and add these following simple commands and run them.

```{r, results="hide"}
a <- 4
b <- 10
a*b
a-b
```

Now, you will notice I use the "<-" symbol to assign a value to a variable.  You can also use "=" and the result is the same.  Why didn't the developers use the equal sign?  The short answer is it originally was used to set function values, for example in the command, rnorm(10, mean = 5), it is setting the mean value to 5 within the rnorm function.  Back in the day there was a single button that made the "<-" and, in fact, in some editors a shortcut still exists.  I will be using the "=" in this tutorial and you can read more about the history [here](http://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html).

There's one last way to assign a number to a variable and this can be very useful if you're creating variables in a loop and, for example, you would like the loop number to be included in the variable name.  
```{r}
assign("a", 5)
a
```