---
title: "R tutorial"
author: "Jeanette Mumford"
output:
  html_document:
    keep_md: yes
    theme: united
    toc: yes
  pdf_document:
    toc: yes
---

## Installing R and getting started
[R](https://www.r-project.org/) is free and I highly recommend using R through the IDE, [RStudio](https://www.rstudio.com/), which offers a lot of helpful functionality like generating R markdown files easily and integrating with github.  A common confusion is that updating RStudio is also updating R, but this is not the case.  RStudio uses R, but they are updated and installed independently.

To install R and RStudio, start by installing R and then install RStudio second.  RStudio should automatically find R when it installs.

Like Python, R requires installing libraries for more advanced tools, but  the installation is typically very easy.  Most libraries you will want to use are located in the [CRAN](https://cran.r-project.org/) (Comprehensive R Archive Network), which can be accessed through R, directly.  The following illustrates how you'd install and load a package.


```{r}
# <-  The comment in R is the hash symbol

# To grab libraries for installation, R will ask you to
# select a CRAN "mirror".  The following line will set the
# mirror permanently so R doesn't ask you over and over again.
#  All the mirrors are listed here: https://cran.r-project.org/mirrors.html

# You only need to run this once ever (not once per session)
options(repos=structure(c(CRAN="http://cran.us.r-project.org")))


# To install a library use the following, note double quotes are used.
# A "newer" feature is dependencies (libraries the library of 
# interest needs) will automatically be installed.

# You only need to run this once ever (not once per session)
install.packages("ggplot2")

# To load the library so you can access the functions 

# Run once per session if you'd like to use it
library("ggplot2")
```

For the above you will likely get more output than I did when you run install.packages, as I already had it installed.

Now open a new editor (in RStudio click the white page icon in the upper left and select, "R Script", and we'll write a quick "hello world".  Enter the following in the file and save it in a file called "first_script.R".
```{r, results = "hide"}
print("Hello world")
```

I feel it is poor coding form to change working directories, so I always recommend typing out the full paths to files.  So edit the following to reflect where you just stored your file.  This will run your script.
```{r}
source("/Users/jeanettemumford/Documents/Research/Teaching/neurohackweek2017/first_script.R")
```

If you want to run the script from the LINUX prompt you would use "Rscript first_script.R".  

In RStudio, you can simply hit cmd+return (on a Mac) to run the current line that the cursor is on, or you can copy and paste into the R terminal window.  Give that a shot with your single line in first_script.R and add these following simple commands and run them.

```{r, results="hide"}
a <- 4
b <- 10
a*b
a-b
```

Now, you will notice I use the "<-" symbol to assign a value to a variable.  You can also use "=" and the result is the same.  Why didn't the developers use the equal sign?  The short answer is it originally was used to set function values, for example in the command, rnorm(10, mean = 5), it is setting the mean value to 5 within the rnorm function.  Back in the day there was a single button that made the "<-" and, in fact, in some editors a shortcut still exists (e.g. Aquamacs).  I prefer fewer keystrokes and don't find it to be confusing, so I will be using the "=" in this tutorial.  You can read a longer answer [here](http://blog.revolutionanalytics.com/2008/12/use-equals-or-arrow-for-assignment.html).

There's one last way to assign a number to a variable and this can be very useful if you're creating variables in a loop and, for example, you would like the loop number to be included in the variable name.  
```{r}
assign("a", 5)
a
```
To insert a variable name, you can combine assign() with another function in R that allows you to combine strings and variables.  The paste0() function and sprintf() allow for this.  
```{r}
x = 10
# paste0 simply connects everything
paste0("variable", x)

# Alternatively, sprintf can work and has some additional flexibilities

# On your own, look at the help for sprintf.  By the way simply type "?" followed
#  by your command name to open the help page
?sprintf
sprintf("variable%s", x)
# the %s acts as a fill in the blank for strings.  You'd need to change this for number input
```
Play around with the paste or sprintf commands.  Try to create a variable called "variable10" that has a value of 4 using the following:
```{r}
a = "variable"
b = 10
c = 4
```

## How data are stored
There are four main data types used in R: numeric, character, factor and logical.  Although numeric, character and logical tend to be more intuitive, factor can often be confusing and sometimes bad things happen if you don't realize something is a factor.  First I'll show each one and then I'll explain what this factor business is.

```{r}
# You can use the class() function to see what type of data R thinks you have

# Numbers (almost) always default to numeric
y = c(1,2,3,4)       # a vector of numbers
class(y)

# Character
names = c("Anne", "Sean", "Lilly", "Logan") # a vector of strings
# Check the class of names()
class(names)

# Factors
gender = c("m", "f", "m", "f", "m","m", "f")  # here is it a character
gender.factor = as.factor(gender)        
gender.factor
class(gender)

# Logicals
a = 1<3
class(a)
# OR
b = c(TRUE, FALSE, TRUE, TRUE)
# Note I didn't use quotes to create b.  There are some special strings in 
# R that don't use quotes, including, NA, NaN, TRUE and FALSE.  That's because these aren't strings, but serve a different purpose.
class(b)
```
The reason R has factors relates to the main purpose of R: run statistical analyses.  In many models you run with categorical variables, there's a specific baseline comparison that makes the most sense.  For example, if I ran a model with three groups, adhd, bipolar and controls, I might prefer my regression coefficients use the control group as the reference and that's exactly what factors do, they allow you to set the reference.  

Probably the place you'll find this the most useful is when using ggplot.  If you want to change the order in which levels are plotted (say in a set of boxplots), releveling a factor into the order you'd like is the solution.  There will be an example later on.

```{r}
group = factor(c("adhd", "bipolar", "ctrl"))
# without doing anything, adhd is listed first, meaning it is the baseline
group

# use relevel to change reference
relevel(group, ref = "ctrl")

```
Sometimes after reading in data it will automatically change things into factors that you wish were numbers.  Be very careful when you do this!  You must first convert to a character and then a numeric, otherwise it will use the level value as the value.  
```{r}
# Common mistake
# Going from factor to number
num.vec = as.factor(c(1.2, 1.4, 1.6))
num.vec

# Wrong way
as.numeric(num.vec)

# Correct way
as.numeric(as.character(num.vec))
```
So far you've only been looking at vectors of numbers, but you'll often want to put a bunch of vectors together and there are multiple options including list, matrix, array, and  data frame.  The first three are found in many software packages, but the data frame is a special structure that makes the data much more useful for a lot of different R functions, including plotting and data manipulation.

The list is best for things of varying lengths.  Function output is typically put in a list.  This is analogous to the MATLAB structure.

Matrices are two dimensional and are best if you're working with an actual matrix.  For example a matrix of fMRI data where the columns correspond to different voxels.  If you plan on using matrix math, you'll want the data to be in a matrix.

Arrays can be multidimensional.

Data frames have columns with different meanings, or variables.  It is basically a list where the lengths of the entries are the same.  Data for input into regression analyses or plotting are typically put into data frames.

```{r}
mat.ex = matrix(c(1,2,3,4,5,6), nrow = 2)
dim(mat.ex)
mat.ex

array.ex = array(c(1:24), c(4, 3, 2))
dim(array.ex)
array.ex

list.ex = list()
list.ex$a = c(1:10)
list.ex$b = c(45:70)
list.ex$c = factor(c("a1", "b1", "c1"))
list.ex$d = as.character(c("bat", "cat", "bird", "dog"))
length(list.ex)
list.ex

col1 = c(1:10)
col2 = c(21:30)
col3 = rep(c("a1", "b1"), each = 5)
data.frame.ex = data.frame(col1, col2, col3)
data.frame.ex
dim(data.frame.ex)
names(data.frame.ex)

# you can easily convert a matrix to a data.frame and add names
df.ex2 = matrix(1:18, ncol = 3)
df.ex2 = data.frame(df.ex2)
names(df.ex2) = c("col1", "col2", "col3")
df.ex2
```

Summary of functions used in this section:
$\beta$

Table Header | Second Header
------------- | -------------
Table Cell | Cell 2
Cell 3 | Cell 4 
class : Tells you the data type in your vector
as.factor/as.character/as.numeric : ways to create factors, characters and numeric data
TRUE, FALSE, NA:  Not actually functions, but special values that R treats as logicals and missing values, respectively
